Perturbing SU($n$)-symmetric interactions

### 2020-01-17 collective_sun/sun_perturbation.tex
added proof that the gap vanishes when \alpha > D

\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}

@@

Here we show that power-law SU($n$)-symmetric interactions of the form
in \eqref{eq:ints} with $h_{pq}=-h/\abs{p-q}^\alpha$ yield a
non-vanishing many-body interaction energy gap when $\alpha<D$, where
$D$ is the dimension of the lattice.  On a periodic lattice of $N=L^D$
spins, the spectral gap of the interaction Hamiltonian is (see
Appendix \ref{sec:trans_inv_single})
\begin{align}
  \Delta_D
  = \sum_{d\in\tilde\ZZ_L^D} h_{0,d} \sp{\cos\p{d \c k_{\t{SE}}}-1}
  = h \sum_{\substack{d\in\tilde\ZZ_L^D\\d\ne0}}
  \f{1-\cos\p{d \c k_{\t{SE}}}}{\abs{d}^\alpha},
\end{align}
where the domain of $d$ is given by
$\tilde\ZZ_L\equiv\set{-\floor{L/2},-\floor{L/2}+1,\cdots,\ceil{L/2}}$,
and $k_{\t{SE}}$ is a wavenumber for a singly-excited spin-wave state.
In order to yield the smallest excitation energy $\Delta_D$, the
wavenumber $k_{\t{SE}}$ should maximize the contribution of the cosine
term above, which is achieved by a wavenumber that minimizes the
oscillations of this term when integrated over the entire lattice.  A
suitable candidate for a minimal wavenumber is
$k_{\t{SE}}=\p{2\pi/L,0,0,\cdots}$, which corresponds to an excitation
energy
\begin{align}
  \Delta_D = h \sum_{\substack{d\in\tilde\ZZ_L^D\\d\ne0}}
  \f{1-\cos\p{d_12\pi/L}}{\abs{d}^\alpha}.
\end{align}
Defining $\epsilon\equiv2/L$ and a rescaled domain
$\SS_\epsilon\equiv\tilde\ZZ_L/\epsilon\subset\sp{-1,1}$, we
substitute $x\equiv\epsilon d\in\SS_\epsilon^D$ to get
\begin{align}
  \Delta_D
  = h \sum_{\substack{x\in\SS_\epsilon^D\\x\ne0}}
  \f{1-\cos\p{\pi x_1}}{\abs{x/\epsilon}^\alpha}
  = h \epsilon^{D-\alpha} \sum_{\substack{x\in\SS_L^D\\x\ne0}}
  \epsilon^D \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha}.
\end{align}
As $\epsilon\to0$ the discrete sum over $x$ becomes an integral that
avoids an infinitesimal region about the origin, i.e.
\begin{align}
  \Delta_D = h \epsilon^{\alpha-D} \I_{D\epsilon},
  &&
  \I_{D\epsilon}
  \equiv \int_{\TT_1^D\setminus\TT_\epsilon^D} \d^Dx\,
  \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha},
\end{align}
where the integral $\I_{D\epsilon}$ is defined using the interval
$\TT_a\equiv\p{-a,a}$.  The integrand of $\I_{D\epsilon}$ is strictly
positive and well-behaved on the entirety of its domain except for the
origin, where depending on the value of $\alpha$ the integrand may
vanish or diverge as $\abs{x}\to0$.  Together, these facts mean that
\begin{align}
  \I_{D\epsilon} \stackrel{\epsilon\to0}{\sim} \epsilon^{-\gamma},
  &&
  \Delta_D \stackrel{\epsilon\to0}{\sim} h \epsilon^{\alpha-D-\gamma},
\end{align}
for some $\gamma\ge0$, which implies that gap $\Delta_D$ is
non-vanishing when
\begin{align}
  \alpha \le D + \gamma,
\end{align}
and in particular when $\alpha \le D$.

>>>>

\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\K}{\mathcal{K}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}

@@

Here we show that power-law SU($n$)-symmetric interactions of the form
in \eqref{eq:ints} with $h_{pq}=-h/\abs{p-q}^\alpha$ yield a
non-vanishing many-body interaction energy gap when $\alpha<D$, where
$D$ is the dimension of the lattice.  On a periodic lattice of $N=L^D$
spins, the spectral gap of the interaction Hamiltonian is (see
Appendix \ref{sec:trans_inv_single})
\begin{align}
  \Delta_D
  = \sum_{d\in\tilde\ZZ_L^D} h_{0,d} \sp{\cos\p{d \c k_{\t{SE}}}-1}
  = h \sum_{\substack{d\in\tilde\ZZ_L^D\\d\ne0}}
  \f{1-\cos\p{d \c k_{\t{SE}}}}{\abs{d}^\alpha},
\end{align}
where the domain of $d$ is given by
$\tilde\ZZ_L\equiv\set{-\floor{L/2},-\floor{L/2}+1,\cdots,\ceil{L/2}}$,
and $k_{\t{SE}}$ is a wavenumber for a singly-excited spin-wave state.
In order to yield the smallest excitation energy $\Delta_D$, the
wavenumber $k_{\t{SE}}$ should maximize the contribution of the cosine
term above, which is achieved by a wavenumber that minimizes the
oscillations of this term when integrated over the entire lattice.  A
suitable candidate for a minimal wavenumber is
$k_{\t{SE}}=\p{2\pi/L,0,0,\cdots}$, which corresponds to an excitation
energy
\begin{align}
  \Delta_D = h \sum_{\substack{d\in\tilde\ZZ_L^D\\d\ne0}}
  \f{1-\cos\p{d_12\pi/L}}{\abs{d}^\alpha}.
\end{align}
Defining $\epsilon\equiv2/L$ and a rescaled domain
$\SS_\epsilon\equiv\tilde\ZZ_L/\epsilon\subset\sp{-1,1}$, we
substitute $x\equiv\epsilon d\in\SS_\epsilon^D$ to get
\begin{align}
  \Delta_D
  = h \sum_{\substack{x\in\SS_\epsilon^D\\x\ne0}}
  \f{1-\cos\p{\pi x_1}}{\abs{x/\epsilon}^\alpha}
  = h \epsilon^{D-\alpha} \sum_{\substack{x\in\SS_L^D\\x\ne0}}
  \epsilon^D \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha}.
\end{align}
As $\epsilon\to0$ the discrete sum over $x$ becomes an integral that
avoids an infinitesimal region about the origin, i.e.
\begin{align}
  \Delta_D = h \epsilon^{\alpha-D} \I_{D\epsilon},
  &&
  \I_{D\epsilon}
  \equiv \int_{\TT_1^D\setminus\TT_\epsilon^D} \d^Dx\,
  \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha},
\end{align}
where the integral $\I_{D\epsilon}$ is defined using the interval
$\TT_a\equiv\p{-a,a}$.  The integrand of $\I_{D\epsilon}$ is strictly
positive and well-behaved on the entirety of its domain except for the
origin, where depending on the value of $\alpha$ the integrand may
vanish or diverge as $\abs{x}\to0$.  Together, these facts mean that
\begin{align}
  \I_{D\epsilon} \stackrel{\epsilon\to0}{\sim} \epsilon^{-\gamma},
  &&
  \Delta_D \stackrel{\epsilon\to0}{\sim} h \epsilon^{\alpha-D-\gamma},
\end{align}
for some $\gamma\ge0$, which implies that gap $\Delta_D$ is
non-vanishing when $\alpha\le D\le D+\gamma$.

For the sake of completion, we add that in fact
$\gamma=\max\set{0,\alpha-D-2}$, which implies that the gap $\Delta_D$
vanishes in the thermodynamic limit when $\alpha>D$.  The value of
$\gamma$ is determined by the behavior of the integrand of
$\I_{D\epsilon}$ near $\abs{x}\sim\epsilon\ll1$, where
$1-\cos\p{\pi x_1}\sim x_1^2$, so
\begin{align}
  \I_{D\epsilon}
  \stackrel{\epsilon\to0}{\sim} \K_{D\epsilon}
  \equiv \int_{\TT_1^D\setminus\TT_\epsilon^D} \d^Dx\,
  \f{x_1^2}{\abs{x}^\alpha}
  = \int_{\TT_1^D\setminus\TT_\epsilon^D} \d y\, \d z^{D-1}
  \f{y^2}{\p{y^2+z^2}^{\alpha/2}},
\end{align}
where for convenience we introduced the variables $y\equiv x_1$ and
$z\equiv\p{x_2,x_3,\cdots,x_D}$.  The integral $\K_{D\epsilon}$ can be
split up into two contributions: a ``bulk'' term
$\K_{D\epsilon}^{\t{B}}$ that integrates over
$\abs{z}\gtrsim\epsilon$, and a remainder $\K_{D\epsilon}^{\t{R}}$
that integrates over $\abs{z}\lesssim\epsilon$.  The domain of $y$ in
each case is determined by the restriction that
$\p{y,z}\notin\TT_\epsilon^D$.  Altogether,
\begin{align}
  \K_{D\epsilon} = \K_{D\epsilon}^{\t{B}} + \K_{D\epsilon}^{\t{R}},
\end{align}
where
\begin{align}
  \K_{D\epsilon}^{\t{B}}
  &\equiv \int_{\TT_1} \d y
  \int_{\TT_1^{D-1}\setminus\TT_\epsilon^{D-1}} \d^{D-1}z\,
  \f{y^2}{\p{y^2+z^2}^{\alpha/2}}, \\
  \K_{D\epsilon}^{\t{R}}
  &\equiv \int_{\TT_1\setminus\TT_\epsilon} \d y
  \int_{\TT_\epsilon^{D-1}} \d^{D-1}z\,
  \f{y^2}{\p{y^2+z^2}^{\alpha/2}}.
\end{align}
Changing to spherical coordinates, we can say that
\begin{align}
  \K_{D\epsilon}^{\t{B}}
  \sim \int_0^1 \d y \int_\epsilon^1 \d z\,
  \f{y^2 z^{D-2}}{\p{y^2+z^2}^{\alpha/2}},
  &&
  \K_{D\epsilon}^{\t{R}}
  \sim \int_\epsilon^1 \d y \int_0^\epsilon \d z\,
  \f{y^2 z^{D-2}}{\p{y^2+z^2}^{\alpha/2}},
\end{align}
where the resulting integrals can be evaluated on a computer algebra
system (e.g.~Mathematica) and expanded to leading orders in $\epsilon$
to find that
\begin{align}
  \K_{D\epsilon}^{\t{B}} \stackrel{\epsilon\to0}{\sim}
  B_0 + B_1 \epsilon^{D-\alpha+2},
  &&
  \K_{D\epsilon}^{\t{R}} \stackrel{\epsilon\to0}{\sim}
  R_1 \epsilon^{D-\alpha+2},
\end{align}
for some constants $B_0,B_1,R_1$ that generally depend on $D$ and
$\alpha$.  It follows that
\begin{align}
  \I_{H\epsilon} \stackrel{\epsilon\to0}{\sim} \K_{D\epsilon}
  = \K_{D\epsilon}^{\t{B}} + \K_{D\epsilon}^{\t{R}}
  \stackrel{\epsilon\to0}{\sim} \epsilon^{-\gamma},
  &&
  \gamma = -\min\set{0,D-\alpha+2}
\end{align}
and so
\begin{align}
  \Delta_D
  \stackrel{\epsilon\to0}{\sim} \epsilon^{\alpha-D-\gamma}
  = \epsilon^{\min\set{\alpha-D,2}}.
\end{align}
